---
title: "README"
format: gfm
---

```{r setup}
#| setup: false
#| echo: false
#| message: false
#| warning: false

library(dplyr)
library(HydeNet)
library(dagitty)
library(ggdag)
library(tidyr)
library(ggplot2)
library(ggthemes)
library(quickdag)
library(entropy)
library(network)
library(ggnetwork)
library(dagitty)
library(ggdag)
library(DiagrammeR)
library(rmutil)
library(jmuOutlier)
library(CausalQueries)

```


This repository contains the Quarto files necessary to reproduce "Holistic Causal Learning with Causal Graphs: A Credible Method for Study Design and Preregistration in the Social Sciences" by Robert Kubinec. Please see the file `code/prereg_causal_graph.qmd` for the main text of the article with embedded calculations of the entropy of causal graphs. The PDF version of the main article with a doi can be found at: https://osf.io/preprints/socarxiv/a492b/.

The appendix, which contains additional analyses, can be found in the document `code/prereg_causal_graph_SI.qmd` in this repository. The appendix contains additional analyses of the entropy of causal graphs.

## Execution and Runtime

These files can be compiled using Rstudio and Quarto or from the command line using only Quarto. See instructions for installing Quarto available here: https://quarto.org/docs/get-started/. To render the files from the command line, use the following commands from the project root directory:

```
quarto render code/prereg_causal_graph.qmd
quarto render code/prereg_causal_graph_SI.qmd
```

From within Rstudio, the file can be executed code chunk by code chunk. The rendering will produce PDFs. Expected execution time should be no more than a few minutes at most.

## System Specification

The files were compiled with R 4.4.1 and make use of the following R packages:

```
library(dplyr)
library(HydeNet)
library(dagitty)
library(ggdag)
library(tidyr)
library(ggplot2)
library(ggthemes)
library(quickdag)
library(entropy)
library(network)
library(ggnetwork)
library(dagitty)
library(ggdag)
library(DiagrammeR)
library(rmutil)
library(jmuOutlier)
library(CausalQueries)
```

Note that two of these packages, `HydeNet` and `quickdag`, are no longer available on CRAN. `HydeNet` can be installed with the `remotes` package (available on CRAN) as follows:

```
remotes::install_github("nutterb/HydeNet")
```

`quickdag` can be installed with the `remotes` package as follows:

```
remotes::install_github("jrgant/quickDAG")
```

The session info from the most recent compilation of the file is as follows:

```{r}
sessionInfo()
```

The file was compiled using Mac OS X Sequoia 15.0 with an M3 chip.